---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# effectsurf

<!-- badges: start -->
[![R-CMD-check](https://github.com/AAGI-AUS/effectsurf/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/AAGI-AUS/effectsurf/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/AAGI-AUS/effectsurf/graph/badge.svg)](https://codecov.io/gh/AAGI-AUS/effectsurf)
<!-- badges: end -->

The goal of {effectsurf} is to extend traditional 2D estimated marginal means (EMMs) into interactive 3D **Estimated Marginal Surfaces** (EMS).
It provides model-agnostic visualisation of predictions, conditional effect sizes, and conditional average treatment effects (CATE) from fitted statistical models.

Surfaces are rendered as interactive [plotly](https://plotly.com/r/) 3D charts with:

- **Categorical stratification** -- overlaid surfaces by group
- **Confidence envelopes** -- upper/lower CI surfaces
- **2D profile projections** -- slices through the 3D surface
- **Derivative surfaces** -- sensitivity, curvature, and interaction maps
- **Post-prediction smoothing** -- smooth jagged surfaces from non-smooth models (e.g., random forests) via `mgcv::gam()` surrogates

Supports 100+ model classes via [marginaleffects](https://marginaleffects.com/), with optional [emmeans](https://cran.r-project.org/package=emmeans) and [gratia](https://gavinsimpson.github.io/gratia/) backends.

## Installation

You can install the development version of {effectsurf} from [GitHub](https://github.com/AAGI-AUS/effectsurf) with:

```{r install, eval = FALSE}
if (!require("pak")) {
  install.packages("pak")
}

pak::pak("AAGI-AUS/effectsurf")
```

## Getting started

{effectsurf} works with any fitted model that supports `predict()`.
Here is a minimal example using a GAM fitted with {mgcv}:

### Fit a model

```{r fit-model}
library(effectsurf)

model <- mgcv::gam(
  mpg ~ s(wt) + s(hp) + factor(cyl),
  data = mtcars
)
```

### Create a prediction surface

```{r prediction-surface}
es <- surf_prediction(
  model,
  x = "wt", y = "hp",
  x_length = 30, y_length = 30,
  labels = list(
    x = "Weight (1000 lbs)",
    y = "Horsepower",
    z = "Fuel efficiency (mpg)"
  )
)
es
```

### Visualise interactively

```{r plot, eval = FALSE}
plot(es)
```

This opens an interactive 3D surface in your browser or RStudio viewer.
Drag to rotate, scroll to zoom, hover for values.

### Stratified surfaces

Overlay surfaces by a categorical variable to visualise interactions:

```{r stratified}
es_strat <- surf_prediction(
  model,
  x = "wt", y = "hp",
  by = "cyl",
  x_length = 25, y_length = 25,
  labels = list(
    x = "Weight (1000 lbs)",
    y = "Horsepower",
    z = "Fuel efficiency (mpg)",
    title = "Prediction surface by cylinder count"
  )
)
es_strat
```

```{r plot-strat, eval = FALSE}
plot(es_strat, opacity = 0.85)
```

### Export to self-contained HTML

```{r export, eval = FALSE}
surf_export(es, "my_surface.html")
```

## Core functions

| Function | Purpose |
|----------|---------|
| `surf_prediction()` | Prediction surface (fitted values across x, y) |
| `surf_comparison()` | Treatment effect surface (contrasts between groups) |
| `surf_slopes()` | Marginal effect surface (partial derivatives) |
| `surf_sensitivity()` | CATE surface (where does treatment work best?) |
| `surf_cate()` | CATE from causal models (e.g., causal forests) |
| `surf_derivatives()` | Gradient, curvature, and interaction maps |
| `surf_profile()` | 2D profile slices from a 3D surface |
| `surf_contour()` | 2D contour plot |
| `surf_optimum()` | Find the (x, y) combination that maximises/minimises z |
| `surf_export()` | Export to self-contained HTML |

All `surf_*()` functions support the `smooth` parameter for post-prediction
surface smoothing via `mgcv::gam()` -- useful for random forests, boosted
trees, and other non-smooth models.

## Agricultural example

{effectsurf} includes a bundled `barley_trials` dataset simulated from
Australian barley trial data (GRDC Project UCS2301-002RTX).

```{r barley}
data(barley_trials)

m_barley <- mgcv::gam(
  yield ~ s(rainfall, k = 5) + nitrogen + seedrate +
    variety_type + state + nitrogen:variety_type +
    s(trial, bs = "re"),
  data = barley_trials,
  method = "REML"
)

es_barley <- surf_prediction(
  m_barley,
  x = "nitrogen", y = "rainfall",
  by = "variety_type",
  x_length = 30, y_length = 30,
  labels = list(
    x = "Nitrogen applied (kg/ha)",
    y = "April-October rainfall (mm)",
    z = "Grain yield (t/ha)",
    title = "Yield response surface by variety type"
  )
)
summary(es_barley)
```

```{r barley-plot, eval = FALSE}
plot(es_barley, opacity = 0.85)
```

## Citation

```{r citation, comment = ""}
citation("effectsurf")
```

## Code of Conduct

Please note that this project is released with a [Contributor Code of
Conduct](https://www.contributor-covenant.org/version/2/1/code_of_conduct/).
By participating in this project you agree to abide by its terms.
