---
title: "Getting Started with effectsurf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with effectsurf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What are Estimated Marginal Surfaces (EMS)?

Traditional Estimated Marginal Means (EMMs) show how a predicted outcome
changes along **one** variable at a time. This is a 1D slice of a
potentially complex response landscape. When two variables interact, these
1D slices can miss critical patterns --- saddle points, interaction ridges,
and optimal management combinations that only become visible in 3D.

**Estimated Marginal Surfaces (EMS)** extend EMMs to **two** continuous
variables simultaneously, producing interactive 3D surfaces. Non-focal
variables are marginalised (held at mean/mode), just as with EMMs.

The `effectsurf` package makes this workflow model-agnostic and requires as
few as **three lines of code**.

## Installation

```{r eval = FALSE}
# Install from local source (during development)
install.packages("effectsurf", repos = NULL, type = "source")
```

## Quick start: three lines to a 3D surface

```{r quick-start}
library(effectsurf)
library(mgcv)

# 1. Fit a model
model <- gam(mpg ~ s(wt) + s(hp) + factor(cyl), data = mtcars)

# 2. Create the prediction surface
es <- surf_prediction(model, x = "wt", y = "hp",
                      x_length = 30, y_length = 30)

# 3. Inspect the object
es
```

To visualise interactively (opens in your browser or RStudio Viewer):

```{r plot-basic, eval = FALSE}
plot(es)
```

## Stratified surfaces: comparing groups in 3D

When you supply a `by` argument, each level of the categorical variable
produces its own surface, overlaid with distinct colours:

```{r stratified}
es_strat <- surf_prediction(
  model, x = "wt", y = "hp", by = "cyl",
  x_length = 25, y_length = 25
)
es_strat
```

```{r plot-strat, eval = FALSE}
plot(es_strat, opacity = 0.85)
```

The colourblind-safe palette (Wong, 2011) ensures accessibility. Each
surface can be toggled on/off via the legend.

## Using the bundled barley trials dataset

`effectsurf` ships with `barley_trials` --- a simulated dataset based on
patterns from Australian national barley agronomy trials:

```{r barley-model}
data(barley_trials)
str(barley_trials)

# Fit a GAM with smooth terms and variety type stratification
barley_model <- gam(
  yield ~ s(rainfall, k = 5) +
    nitrogen + seedrate + variety_type + state +
    nitrogen:variety_type +
    s(trial, bs = "re"),
  data = barley_trials,
  method = "REML"
)

summary(barley_model)$r.sq
```

```{r barley-surface}
# Create yield response surface: nitrogen x rainfall, by variety type
es_barley <- surf_prediction(
  barley_model,
  x = "nitrogen", y = "rainfall",
  by = "variety_type",
  x_length = 30, y_length = 30,
  labels = list(
    x = "Nitrogen applied (kg/ha)",
    y = "April-October rainfall (mm)",
    z = "Grain yield (t/ha)",
    title = "Yield response by variety type"
  )
)
es_barley
```

```{r barley-plot, eval = FALSE}
plot(es_barley, opacity = 0.85)
```

## 2D profile projections

Extract publication-ready 2D slices from the 3D surface using
`surf_profile()`:

```{r profiles, fig.width = 7, fig.height = 4}
# How does yield change with nitrogen at different rainfall levels?
surf_profile(es_barley, along = "x", at = c(150, 300, 500))
```

## Finding the surface optimum

```{r optimum}
surf_optimum(es_barley, type = "max")
```

## Exporting to HTML

Save a self-contained interactive HTML file for sharing:

```{r export, eval = FALSE}
surf_export(es_barley, path = "barley_yield_surface.html")
```

## Summary of key functions

| Function | Purpose |
|----------|---------|
| `surf_prediction()` | 3D prediction surface |
| `surf_comparison()` | Treatment effect difference surface |
| `surf_slopes()` | Marginal effect (derivative) surface |
| `surf_cate()` | CATE surface for causal models |
| `surf_profile()` | 2D profile projections |
| `surf_contour()` | 2D contour/heatmap |
| `surf_sensitivity()` | Treatment effect landscape |
| `surf_optimum()` | Surface maximum/minimum |
| `surf_compare()` | Model comparison (difference) surface |
| `surf_export()` | Export to self-contained HTML |

## Next steps

- **Vignette 2**: Stratified Surfaces --- comparing groups in 3D
- **Vignette 3**: Treatment Effects in 3D --- from effect sizes to CATE
- **Vignette 4**: Model-Agnostic Surfaces --- GAMs, mixed models, and ML
